Do researchers preferentially collaborate with colleagues of the same gender?

### 2018-06-03 manuscript/Manuscript_Claire.Rmd
Luke writing

Women are highly underrepresented in many branches of science, technology, engineering, mathematics, and medicine (STEMM), and face additional challenges and inequities relative to men [e.g. @Shaw_2012; @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions [@Wutte_2007; @Reuben_2014] with lower salaries [@Trower_2002; @Umbach_2007], receive less grant money [@Hosek_2005; @OER_2008], are promoted more slowly [@Zuckerman_1987; @Rosenfeld_1991; @Long_1993; @Hopkins_2013], and are allocated fewer resources [@ODorchai_2009] and less research funding [@Feldt_1986; @Stack_2004; @Lariviere_2011] than men. Experimental studies suggest that women's research achievements are regarded less favourably than identical achievements by men [@Moss_2012; @Knobloch_2013].

Because publishing, networking and collaboration are instrumental to scientific productivity and academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], dozens of studies have tested for gender differences in these areas [e.g. @Long_1992; @Bozeman_2011; @Abramo_2013; @Badar_2013; references in Table S1 of @Holman_2018]. For example, studies have found that women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less within their own university departments [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in total [@Fuchs_2001]. These gender differences in collaboration practice presumably have multiple causes, which might include implicit and explicit gender bias [@Moss_2012], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], concerns relating to sexual harassment [@jagsi_2016], and unequal access to conferences [@Martin_2014] and travel funds [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008; @Cohn_2014]. Additionally, empirical studies imply that mixed-gender [@Bear_2011; @Campbell_2013] or otherwise diverse [@Hong_2004] teams produce better results on collaborative tasks than less diverse teams. For reasons such as these, several studies have tested for gender differences in collaboration frequency or pattern by examining the author lists of published research. To our knowledge, every published study of this question suggests that men co-publish with other men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing has been termed 'gender homophily'.

However, we believe that prior studies of gender homophily were hindered by a largely unacknowledged statistical issue that we term the Wahlund effect (\autoref{wahlund_plot}), by analogy with a conceptually similar effect in population genetics [@Wahlund_1928]. The Wahlund effect makes it deceptively difficult to infer gender-based preferences simply by counting same- and mixed-gender coauthorships. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if opposite-gendered colleagues are prefered. For example, a sample of literature containing bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018].

@@

In the present study, we test whether researchers working in the life sciences tend to co-publish with same-gendered collaborators, while controlling for the Wahlund effect as strictly as possible. Our study uses a recently-published dataset describing the gender of 35.5m authors from 9.15m articles indexed on PubMed [@Holman_2018]. Holman et al. [-@Holman_2018] reported large differences in the gender ratio of authors across research disciplines, journals, countries, and across the years 2002-2016. We therefore tested for gender homophily after restricting the analysis to particular journals (i.e. research specialties), time periods, and countries.

@@

We also found that gender homophily was marginally stronger in 2015-2016 relative to 2005-2006. Although this trend might reflect a change in the gender preferences of researchers seeking collaborators over time, there are alternative (and perhaps more likely) explanations. For example, this trend might result from the increasing number of women working in senior positions in STEMM over the past decade [e.g. @Long_2015; @Bendels_2018; @McKenzie_2017]. As shown in \autoref{simulation}, if enough coauthorships are between junior and senior researchers, a large gender gap between career stages can give the appearance of heterophily. As this gender gap between career stages lessens, the observed values of $\alpha$ may increase.

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since Sheltzer and Smith [-@sheltzer_2014]'s results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or the perception that these groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], this result provides non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Campbell_2013].

>>>>

Women are severely underrepresented in many branches of science, technology, engineering, mathematics, and medicine (STEMM), and face additional challenges and inequities relative to men [e.g. @Shaw_2012; @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions [@Wutte_2007; @Reuben_2014] with lower salaries [@Trower_2002; @Umbach_2007], receive less grant money [@Hosek_2005; @OER_2008], are promoted more slowly [@Zuckerman_1987; @Rosenfeld_1991; @Long_1993; @Hopkins_2013], and are allocated fewer resources [@ODorchai_2009] and less research funding [@Feldt_1986; @Stack_2004; @Lariviere_2011]. Experimental evidence suggests that bias against women plays a major role in generating these differences [@Moss_2012; @Knobloch_2013].

Because publishing, networking and collaboration are instrumental to scientific productivity and academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], dozens of studies have tested for gender differences in these areas [e.g. @Long_1992; @Bozeman_2011; @Abramo_2013; @Badar_2013; references in Table S1 of @Holman_2018]. For example, studies have found that women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less within their own university departments [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in total [@Fuchs_2001]. These gender differences in collaboration practice presumably have multiple causes, which might include implicit and explicit gender bias [@Moss_2012], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], gender differences in confidence or self-esteem [@bleidorn2016age], concerns relating to sexual harassment [@jagsi_2016], and unequal access to conferences [@Martin_2014] and travel funds [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008]. Additionally, empirical studies imply that mixed-gender or otherwise diverse teams produce better results on collaborative tasks than less diverse teams [@reagans2001; @Hong_2004; @Bear_2011; @Campbell_2013]. For reasons such as these, several studies have examined the author lists of published research articles in order to test for gender differences in collaboration frequency or pattern. To our knowledge, all of these studies imply that men co-publish with other men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing is often termed 'gender homophily'.

However, we believe that prior studies of gender homophily were hindered by a largely unacknowledged statistical issue that we name the Wahlund effect (\autoref{wahlund_plot}), by analogy with the conceptually similar Whalund effect in population genetics [@Wahlund_1928]. The Wahlund effect makes it deceptively difficult to infer gender-based preferences simply by counting the number of same- and mixed-gender coauthorships. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if opposite-gendered colleagues are preferred. For example, a sample of literature containing bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018].

@@

In the present study, we test whether life sciences researchers tend to co-publish with same-gendered colleagues, while controlling for the Wahlund effect as strictly as possible. We use a recently-published dataset describing the gender of 35.5m authors from 9.15m articles indexed on PubMed [@Holman_2018]. Holman et al. [-@Holman_2018] reported large differences in the gender ratio of authors across research disciplines, journals, countries, and across the years 2002-2016. We therefore tested for gender homophily after restricting each analysis to particular journals (i.e. research specialties), time periods, and countries.

@@

We also found that gender homophily was marginally stronger in 2015-2016 relative to 2005-2006. Although this trend might reflect a change in the gender preferences of researchers seeking collaborators over time, there are alternative (and perhaps more likely) explanations. For example, this trend might result from the increasing number of women working in senior positions in STEMM over the past decade [e.g. @Shaw_2012; @Long_2015; @McKenzie_2017; @Bendels_2018]. As shown in \autoref{simulation}, if enough coauthorships are between junior and senior researchers, a large gender gap between career stages can give the appearance of heterophily. As this gender gap between career stages lessens, the observed values of $\alpha$ may increase.

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since Sheltzer and Smith [-@sheltzer_2014]'s results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or the perception that these groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], this result provides non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@reagans2001; @Hong_2004; @Bear_2011; @Campbell_2013].
